A Bidirected String Graph Model for Genome Assembly

### 2012-12-07 presentation/paper.tex
paper: bidiedge interp

The solution to all these problems is to use a new type of graph called a
\KeyTerm{bidirected graph}:
\begin{center}
	\begin{tikzpicture}[>=triangle 45]
		\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
		\node (f) at (0, 0) {f};
		\node (g) at (1.7, 0) {g};
		\node (h) at (3.4, 0) {h};
		\draw[>->,style=very thick] (f) edge (g);
		\draw[>-<,style=very thick] (g) edge (h);
		\draw[>-<,style=very thick,bend left] (f) edge (h);
	\end{tikzpicture}
\end{center}

@@

@@

>>>>

The solution to all these problems is to use a new type of graph called a
\KeyTerm{bidirected graph}.
\begin{center}
	\begin{tikzpicture}[>=triangle 45]
		\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
		\node (f) at (0, 0) {f};
		\node (g) at (1.7, 0) {g};
		\node (h) at (3.4, 0) {h};
		\draw[>->,style=very thick] (f) edge (g);
		\draw[>-<,style=very thick] (g) edge (h);
		\draw[>-<,style=very thick,bend left] (f) edge (h);
	\end{tikzpicture}
\end{center}

@@

The concept of a walk through a bidirected graph maps quite nicely onto the
genome assembly problem, since a walk through the bidirected string graph
constructed from reads and overlaps models a way in which the reads can be
assembled together consistently.  In such a walk, each read may be used in
forward or reverse-complement orientation, depending on the direction, relative
to the walk, of the adjacent arrow heads when the corresponding vertex is passed
through.

For example, consider the walk $1 \to 2 \to 3 \to 4$ in
\Figure{fig:bidigraph_mapping}.  When the walk passes through vertex 1, the
adjacent arrow head is pointed opposite the direction of the walk.  This implies
that the corresponding read is used in reverse-complement orientation.  When the
walk passes through vertex 2, the adjacent arrow heads are pointed in the
direction of the walk, which implies that the corresponding read is used in
forward orientation.  Finally, when the walk passes through vertices 3 and 4,
the adjacent arrow heads are pointed opposite the direction of the walk, which
implies that the corresponding reads are used in reverse-complement orientation.
Note that $4 \to 3 \to 2 \to 1$ is also a valid walk, which represents the way
in which the four reads could also assemble the other strand of the DNA
molecule.

\begin{figure}
	\begin{center}
		{\small
		\begin{tikzpicture}[>=triangle 45]
			\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
			\node (1) at (0.00, 0) {1};
			\node (2) at (2.0, 0) {2};
			\node (3) at (4.0, 0) {3};
			\node (4) at (6.0, 0) {4};
			\draw[<->,very thick] (1) edge (2);
			\draw[>-<,very thick] (2) edge (3);
			\draw[<-<,very thick] (3) edge (4);
		\end{tikzpicture} }

		\vspace{2.5mm}

		\begin{tikzpicture}[>=\ReadArrowType]
				\draw[<-,style=very thick] (0.0, 0.0) -- (2.6, 0.0);
				\draw[->,style=very thick] (1.6, -0.3) -- (4.2, -0.3);
				\draw[<-,style=very thick] (3.2, -0.6) -- (5.8, -0.6);
				\draw[<-,style=very thick] (4.8, -0.9) -- (7.4, -0.9);
				\draw (-0.3, 0.0) node {1.};
				\draw (-0.3, -0.3) node {2.};
				\draw (-0.3, -0.6) node {3.};
				\draw (-0.3, -0.9) node {4.};
		\end{tikzpicture}
	\end{center}
	\caption{A bidirected graph and the implied assembly of the
	corresponding reads}
	\label{fig:bidigraph_mapping}
\end{figure}

Earlier, a crucial point was quickly skipped over.  Each edge is supposed to be
labeled with a DNA sequence, as we are building a bidirected {\em string} graph.
However, we actually need to label each edge with two DNA sequences: one for
each direction in which the edge can be traversed.  To see this, we need to
consider exactly what it means to walk between vertices in the bidirected graph.

Consider the the string graph fragment shown in \Figure{fig:bidiedge_interp_a}.
When we walk from $f \to g$, we are using both $f$ and $g$ in forward
orientation, based on the arrow heads.  But what this really means is that we
are walking from the {\em end} of read $f$ to the {\em end} of read $g$.  In
other words, traversing the edge in that direction is associated with the
addition of the sequence after the overlapped region in $g$, up to the end of
the $g$.

If we instead walk $g \to f$, we instead use both $g$ and $f$ in
reverse-complement orientation by walking from the {\em beginning} of read $g$
to the {\em beginning} of read $f$.  In other words, traversing the edge in that
direction is associated with the addition of the sequence before the overlapped
region in $f$, up to the beginning of $f$, and this sequence must be
reverse-complemented because it is being used opposite its original direction.

These interpretations can be generalized to the other two types of bidirected
edges.  Furthermore, these intepretations imply that the actual sequence spelled
out by a walk in the bidirected graph may include implied (non-labeled) sequence
at the start (if we start at the end of a read) and/or the end (if we end at the
beginning of a read).

\begin{figure}
	\centering
	\begin{subfigure}{0.48\textwidth}
		\begin{tikzpicture}[>=triangle 45,scale=1.5]
			\node[circle,fill=blue!20,draw=black] (f) at (0, 0) {f};
			\node[circle,fill=blue!20,draw=black] (g) at (2.0, 0) {g};
			\draw[>->,style=very thick] (f) edge (g);
			\draw (1.00, 0.1) node[anchor=south] {\small \tt TTAGAAGTA};
			\draw (1.00, -0.1) node[anchor=north] {
					\rotatebox{180}{\small \tt
					TGTGTAATC}};
			\draw[->] (0.5, 0.5) ..
				controls (0.7, 0.7) and (1.3, 0.7) ..
				(1.5, 0.5);
			\draw[->] (1.5, -0.5) ..
				controls (1.3, -0.7) and (0.7, -0.7) ..
				(0.5, -0.5);
		\end{tikzpicture}
		\caption{bidirected string graph fragment}
		\label{fig:bidiedge_interp_a}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\begin{tikzpicture}[>=\ReadArrowType,scale=0.8]
			\draw[->,very thick] (0, 0) node[anchor=east] {$f.B$}
					-> (4, 0) node[anchor=west] {$f.E$};
			\draw[->,very thick] (2.09, -0.9) node[anchor=east] {$g.B$}
					-> (6.09, -0.9) node[anchor=west] {$g.E$};

			\draw (0, -0.1) node[anchor=south west]
				{\hspace*{-1.4mm}{\small \tt
					%GATTACACAGTGTAGTT
					GATTACACA\hspace*{0.1cm}$a_1 \dots a_n$
				}};

			\draw (0, 0.1) node[anchor=north west]
				{\hspace*{-2.6mm}
					\rotatebox{180}{{\small \tt TGTGTAATC}}};

			\draw (2.09, -1.0) node[anchor=south west]
				{\hspace*{-1.4mm}{\small \tt
					%GTGTAGTTTTAGAAGTA
					\hspace*{0.1cm}$a_1 \dots a_n$\hspace*{0.25cm}TTAGAAGTA
				}};

			\draw (2.09, -0.9) node[anchor=north west]
				{\hspace*{12.48mm}
					\rotatebox{180}{{\small \tt TACTTCTAA}}};

			\draw (2.09, -1.4) node[anchor=north] {$o.g_{begin}$}
					-- (2.09, 0.6) node[anchor=south] {$o.f_{begin}$};
			\draw (3.99, -1.4) node[anchor=north] {$o.g_{end}$}
					-- (3.99, 0.6) node[anchor=south] {$o.f_{end}$};
		\end{tikzpicture}
		\caption{The corresponding overlap (labeled)}
		\label{fig:bidiedge_interp_b}
	\end{subfigure}

	\begin{subfigure}{0.5\textwidth}
		\begin{tabular}{c|c}
			walk & interpretation \\ \hline
			$f \to g$ & $f.E \to g.E$ with sequence $g[o.g_{end} + 1 \dots ${\tt
			length}$(g)]$ \\
			$g \to f$ & $g.B \to f.B$ with sequence $f[o.g_{begin} - 1 \dots 0]$ \\
		\end{tabular}
		\caption{Interpretation of each walk on the graph fragment}
	\end{subfigure}
	\caption{Full interpetation of an edge in the bidirected string graph}
	\label{bidiedge_interp}
\end{figure}

@@

There is yet another flaw in the way that the construction of the graph has been
presented that needs to be addressed.  Earlier, the definition of a
\KeyTerm{contained read} was provided.
